Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-03 doc/papers/cgo2014/paper.tex
CGO: review

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However as
    these languages favor a high-level coding style, the computation kernels are
    different from what we use to see in traditional C or Fortran programs:
    explicit raw loops are replaced by high-level constructs and calls to the
    relevant toolbox/module/library/package. As a consequence, traditional
    loop optimization techniques like fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

@@

\keywords
static compilation, parallelization, Python, C++, SIMD, runtime

@@

There is no need to reinvent the wheel when compiling numerical Python kernels.
A profitable approach is to map Python high-level to existing ones, or
translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
As a consequence, this section focuses on
the translation of Python constructs to efficient C++ constructs. The choice of
C++ instead of C is motivated by the opportunities offered by meta programming
and the interesting similarities between C++11~\cite{isocxx11} features and
Python constructs.

@@

A naïve translation of such call to a generic C++ function would generate a
call to a C++ function that would allocate enough memory to hold three integer,
fill the corresponding buffer and return it. A smarter version would generate a
static array holding the relevant values. Figure~\ref{fig:unroll-range}
compares two translation of a Python snippet that iterates over a fixed-size
list. The Clang compiler successfully turns Listing~\ref{lst:stat} into a
single constant. It also statically computes the return value for
Listing~\ref{lst:dyna}, but still generates the initialization code for the
\texttt{std::vector}. The GCC compiler fails to statically compute the return
value in both cases. This example shows that constant propagation should be
done at a higher level than C++, and so be part of the Python optimizer.

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its arguments or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially easy in Python thanks to the
ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

Once inter procedural constant evaluation is done (eventually followed by
forward substitution), iteration over fixed size lists may appear, giving an
opportunity to fully unroll these loops at the Python level, thus avoiding the
suboptimal C++ translation observed for Listing~\ref{lst:stat}
and~\ref{lst:dyna}.

@@

A simplified version of the evaluation loop for an expression template is
illustrated on Listing~\ref{lst:eval-et}. The loop over each dimension of the
target array expression \texttt{to} is made explicit through a recursive
template call. The computation of the origin expression is only triggered for
the innermost loop.

@@

A naive way to parallelize the parametric loop nest from
Listing~\ref{lst:eval-et} is to use a \texttt{parallel} directive from the
OpenMP~\cite{openmp4} standard. Depending on the number of dimension of the
input array, this may lead to nested parallel regions. When OpenMP nested
parallel regions are activated, each parallel region spawns as many thread as
possible up to the runtime upper bound, which leads to oversubscription with
more common implementation (libgomp and libiomp) and generally bad performances. When
disabled, only the first parallel region spawns threads, which is likely to be
inefficient depending on the input array shape, e.g. $(1,100000)$. Moreover,
even if they do not start new threads, each nested parallel region implies a
costly overhead for scheduling computation and data sharing. An easy way to solve
the issue with small then large dimension is to use the \texttt{collapse}
clause but it is not compatible with the parametric loop version.

@@

A conservative approach is to only start a parallel region if the loop trip
count is large enough. We avoid by using the \texttt{if} clause for parallel
regions to do so in order to prevent the unnecessary overhead. Instead a manual
if is used: the loop statement is duplicated and guarded by the appropriate
check, where only one loop holds the parallel annotations.

Note that this parallelization scheme is compatible with the manual
parallelization of Python programs proposed in~\cite{pyhpc2013}.

@@

A first exploration of vectorization of Python constructs is done
in~\cite{wpmvp2014}. It shows that language constructs like list comprehension
offer many vectorization opportunities. This section focuses on the
vectorization of complex Numpy expressions. In order to vectorize a numpy
expression, the last dimension of the expression must be contiguous. This can
be checked at compile time using type traits. The general form of a subscripted
array can be expressed in C++ using a template type, say \texttt{numpy\_gexpr}.
For instance the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix is
represented by the C++ type in Listing~\ref{lst:gexpr}.

\begin{lstlisting}[language=c++, caption={C++ type corresponding to the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix}, label={lst:gexpr}]
numpy_gexpr<ndarray<double, 3>,
            contiguous_slice,
            long, slice>
\end{lstlisting}

This representation makes it easy to
statically compute, in a vectorizable type trait, wether the last dimension of
a subscript is contiguous or not. It can be generalized to the other array
manipulations. BOOST.simd~\cite{esterie2012boost} provides a
machine-independant abstraction for vector operation and supports all the C99
math library. Thanks to this abstraction, all Numpy binary and unary functions
and operators can be vectorized.

@@

As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
to evaluate the optimizations described in\S~\ref{sec:pythran}. It uses Pythran (commit: 488b7ba)
with optimization enabled/disabled. The results are compared with Pypy \todo{add}(???),
Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.

@@

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We refer to this configuration as \textbf{AMD48}.
On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
from OpenMP improvement. This clang version is based on clang 3.4 release
(LLVM:233b1e3f034, clang-omp:6f358bcd87).

%FIXME: je me demande si on devrait pas utiliser juste cette archi. Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...
The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 processors. This configuration
provides a total of 63\,GB of main memory. It supports up to AVX. We
refer to this configuration as \textbf{Intel8}.
On Intel8, backend compilers is gcc 4.9 (20140528).

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes, we run the numpy-benchmarks test
suite multiple time on the \textit{rosen} benchmark.
Figures~\ref{fig:rosenamd48} and~\ref{fig:rosenintel8} illustrate the results.
The difference between the two architectures can be explain by AVX support on
the Intel8 architecture while AMD48 supports up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
% tODO : I have to redo mesures without forward substitution to show difference.
% SG: may be the number of iteration is not enough? what array size did you use? (it should be specified in the text secion, BTW

@@

Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
use of the techniques described in this paper. As we can see, both SIMD instruction
and parallelization have an impact on execution time. We can notice that SIMD
contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
Also, we can notice that gcc as a backend compiler finds more optimization opportunities.
%FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
% ça allègera aussi les graphes...

@@

At the opposite, Figure~\ref{fig:diffusion} shows that clang is better when generated
code make a more intensive use of templates (heavily used with \texttt{gsliced\_ndarray}).
\texttt{gsliced\_ndarray} is a non aligned view on an array in this case so SIMD optimization
can't be applied on the second dimension even if loop fusion and temporary
elimination is performed.  We can notice that Parakeet performs really well on
extensive gslice use.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
%SG: +1

@@

Finally, Figure~\ref{fig:l2norme} show the l2norme algorithm introduce in
\S~\ref{sec:stackoverflow}. Comparison with other compiler show that Pythran approach
is correct for these kind of functions.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
%SG: +1
\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
\caption{Execution time of the l2norm kernel on AMD48.}
\label{fig:l2norme}
\end{figure}

>>>>

    As more scientists use these scripting languages, there is an on-going
    research effort to compile them into efficient native code. However as
    these languages favor a high-level coding style, the computation kernels are
    different from what we use to see in traditional C or Fortran programs:
    explicit raw loops are replaced by high-level constructs and calls to the
    relevant toolbox/module/library/package. As a consequence, traditional
    algorithms to perform
    loop optimization techniques like fusion, tiling and related
    vectorization or parallelization techniques are not directly applicable.

@@

\keywords
Static Compilation, Parallelization, Python, C++, SIMD, Runtime

@@

There is no need to reinvent the wheel when compiling numerical Python kernels.
A profitable approach is to map Python high-level to existing ones, or
translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
As a consequence, this section focuses on the translation of Python constructs
to efficient C++ constructs. The choice of
C++ instead of C is motivated by the opportunities offered by meta programming
and the interesting similarities between C++11~\cite{isocxx11} features and
Python constructs.
The meta-programming part of C++ makes it more than a traditional compiler
runtime, closer to a code generator backend.

@@

A naive translation of such call to a generic C++ function would generate a
call to a C++ function that would allocate enough memory to hold three integers,
fill the corresponding buffer and return it. A smarter version would generate a
static array holding the relevant values. Figure~\ref{fig:unroll-range}
compares two translations of a Python snippet that iterates over a fixed-size
list. The Clang compiler successfully turns Listing~\ref{lst:stat} into a
single constant. It also statically computes the return value for
Listing~\ref{lst:dyna}, but still generates the initialization code for the
\texttt{std::vector}. The GCC compiler fails to statically compute the return
value in both cases. This example shows that constant propagation should be
done at a higher level than C++, and so be part of the Python optimizer.

The argument effect analysis described in \S~\ref{sec:fs} can be extended to
global variables to determine if a function has side effects on its arguments or
on global variables (including I/O). This information makes it possible to flag
some function as pure, and to statically evaluate pure function calls with
constant arguments. This evaluation is especially feasible in Python thanks to
the ability of the language to turn an AST expression into bytecode, evaluate it
and retrieve its constant value. For instance, this makes it possible to turn a
call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
value of \texttt{fibonnaci(4)} at compile-time.

@@

Once interprocedural constant evaluation is done (eventually followed by
forward substitution), iteration over fixed size lists may be exposed, giving an
opportunity to fully unroll these loops at the Python level, thus avoiding the
suboptimal C++ translation observed for Listing~\ref{lst:stat}
and~\ref{lst:dyna}.

@@

A simplified version of the evaluation loop for an expression template is
illustrated on Listing~\ref{lst:eval-et}. The loop over each dimension of the
target array expression \texttt{to} is made explicit through a recursive
template call. The computation of the \texttt{from} expression is only triggered
for the innermost loop.

@@

A naive way to parallelize the parametric loop nest from
Listing~\ref{lst:eval-et} is to use a \texttt{parallel} directive from the
OpenMP~\cite{openmp4} standard. Depending on the number of dimensions of the
input array, this may lead to nested parallel regions. When OpenMP nested
parallel regions are activated, each parallel region spawns as many threads as
possible up to the runtime upper bound, which leads to oversubscription with
more common implementation (libgomp and libiomp) and generally bad performances. When
disabled, only the first parallel region spawns threads, which is likely to be
inefficient depending on the input array shape, e.g. $(1,100000)$. Moreover,
even if they do not start new threads, each nested parallel region implies a
costly overhead for scheduling computation and data sharing. An easy way to solve
the issue with small then large dimension is to use the \texttt{collapse}
clause but it is not compatible with the parametric loop version.

@@

A conservative approach is to only start a parallel region if the loop trip
count is large enough. We avoid by using the \texttt{if} clause for parallel
regions to do so in order to prevent the unnecessary overhead. Instead a manual
if is used: the loop statement is duplicated and guarded by the appropriate
check, where only one loop holds the parallel annotations.
\todo{this paragraph seems to be about a particular runtime issue, this shouldn't matter.
Moreover, how is it relevant to this paper? It is an "implementation detail"}

Note that this parallelization scheme is compatible with the manual
parallelization of Python programs proposed in~\cite{pyhpc2013}.
\todo{I think a specific section could be made on this topic}

@@

A first exploration of vectorization of Python constructs is done
in~\cite{wpmvp2014}. It shows that language constructs like list comprehension
offer many vectorization opportunities. This section focuses on the
vectorization of complex Numpy expressions. In order to vectorize a Numpy
expression, the last dimension of the expression must be contiguous. This can
be checked at compile time using type traits. The general form of a subscripted
array can be expressed in C++ using a template type, say \texttt{numpy\_gexpr}.
For instance the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix is
represented by the C++ type in Listing~\ref{lst:gexpr}.

\begin{lstlisting}[language=c++, caption={C++ type corresponding to the Numpy subscript \lstinline{a[:,0,::2]} on a 3D float matrix}, label={lst:gexpr}]
numpy_gexpr<ndarray<double, 3>,
            contiguous_slice,
            long, slice>
\end{lstlisting}

This representation makes it easy to
statically compute, in a vectorizable type trait, whether the last dimension of
a subscript is contiguous or not. It can be generalized to the other array
manipulations. BOOST.simd~\cite{esterie2012boost} provides a
machine-independant abstraction for vector operation and supports all the C99
math library. This abstraction helps to vectorize all Numpy binary and unary
functions and operators in Pythran's runtime.

@@

As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test 
suite to evaluate the optimizations described in\S~\ref{sec:pythran}. 
Pythran is used (commit: 488b7ba) with and without SIMD optimization. 
The results are compared with Pypy \todo{add}(???), Parakeet (commit: 4e55810), 
Numba (commit: b8a79c5) and Python 2.7.

@@

The first one holds 8 AMD Magny Cours processors for a
total of 48 cores. Each core has access to 64\,KB of L1 cache,
512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 6 cores of a processor. This configuration
provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
up to SSE4. We refer to this configuration as \textbf{AMD48}.
On AMD48, the backend compilers are GCC (Debian 4.7.2-5) and the clang-omp
version from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to
benefits from OpenMP improvements. This Clang version is based on Clang 3.4
release (LLVM:233b1e3f034, clang-omp:6f358bcd87).

\todo{je me demande si on devrait pas utiliser juste cette archi. 
Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...}
\todo{Mehdi: agree, sauf si on peut avoir UN exemple avec 48 threads qui scale 
"correctement"}

The second one holds one Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
8 cores (multithreading is not used). Each node can access to 64\,KB of L1
cache, 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
is shared between the 8 cores. This configuration provides a total of 64\,GB of 
main memory. It supports up to AVX. We refer to this configuration as
\textbf{Intel8}.
On Intel8, backend compilers is GCC 4.9 (20140528).

@@

% For performance comparison between different Pythran version
To compare the different optimization schemes, we run the numpy-benchmarks test
suite multiple times on the \textit{rosen} benchmark.
Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
The difference between the two architectures can be explain by AVX support on
the Intel8 architecture while AMD48 supports up to SSE4 only.
% Haaaaa, I don't understand why OMP is not use...
\todo{I have to redo mesures without forward substitution to show difference.}
% SG: may be the number of iteration is not enough? what array size did you use? (it should be specified in the text secion, BTW

@@

Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
use of the techniques described in this paper. As we can see, both vectorization
and parallelization have an impact on execution time. We can notice that SIMD
contribution is not as good on AMD48 as it is on Intel8, because it does not
support AVX.
%FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
% ça allègera aussi les graphes...

@@

Figure~\ref{fig:diffusion} show how Clang generated code is more performant
when handling \lstinline{gsliced_ndarray}.
This structure is a non aligned view on an array which prevent SIMD optimization
on the second dimension even if loop fusion and temporary elimination is
performed.  We can notice that Parakeet performs really well on
extensive gslice use.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
%SG: +1

@@

Finally, Figure~\ref{fig:l2norme} shows the L2norm algorithm introduced in
\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
approach taken by Pythran is relevant for these kind of functions.
%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
%SG: +1
\begin{figure}[t]
\centering
\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
\caption{Execution time of the l2norm kernel on AMD48.}
\label{fig:l2norme}
\end{figure}
